%!TEX root = ../thesis.tex
% ******************************* Thesis Appendix C ********************************

\chapter{Publications contributed to during the PhD degree} \label{appendix:pubs}
This Appendix lists the publications to which I contributed as part of work developed during my PhD research. List updated at time of submission.\newline

Vieira Braga F, Kar G, Berg M, Carpaij O, Polanski K, Simon L, Brouwer S, \textbf{Gomes T}, Hesse L, Jiang J, Fasouli E, Efremova M, Vento-Tormo R, Talavera-López C, Jonker M, Affleck K, Palit S, Strzelecka P, Firth H, … Teichman, SA. (2019) A cellular census of human lungs identifies novel cell states in health and in asthma. Nature Medicine 25: 1153-1163\newline

Miragaia, R.*, \textbf{Gomes, T.*}, Chomka, A., Jardine, L., Riedel, A., Hegazy, A., Whibley, N., Tucci, A., Chen, X., Lindeman, I., Emerton G, Krausgruber T, Shields J, Haniffa M, Powrie F, and Teichmann S. (2019) Single-Cell Transcriptomics of Regulatory T Cells RevealsTrajectories of Tissue Adaptation. Immunity 50, 493-504.e7.\newline

Lun A, Riesenfeld S, Andrews T, Dao T, \textbf{Gomes T}, and Marioni J (2019) EmptyDrops: distinguishing cells from empty droplets in droplet-based single-cell RNA sequencing data. Genome Biology 20:\newline

Henriksson J, Chen X, \textbf{Gomes T}, Ullah U, Meyer K, Miragaia R, Duddy G, Pramanik J, Yusa K, Lahesmaa R, and Teichmann SA. (2019) Genome-wide CRISPR Screens in T Helper Cells Reveal Pervasive Crosstalk between Activation and Differentiation. Cell 176: 882-896.e18\newline

Hagai T, Chen X, Miragaia R, Rostom R, \textbf{Gomes T}, Kunowska N, Henriksson J, Park J, Proserpio V, Donati G, Bossini-Castillo L, Vieira Braga F, Naamati G, Fletcher J, Stephenson E, Vegh P, Trynka G, Kondova I, Dennis M, … Teichmann, SA. (2018) Gene expression variability across cells and species shapes innate immunity. Nature 563: 197-202\newline

Kunz, DJ; \textbf{Gomes, T}; James, KR; Immune cell dynamics unfolded by single-cell technologies, (2018), Frontiers in immunology, 9, 1435\newline

Pramanik J, Chen X, Kar G, Henriksson J, \textbf{Gomes T}, Park J, Natarajan K, Meyer K, Miao Z, McKenzie A, Mahata B, and Teichmann S (2018) Genome-wide analyses reveal the IRE1a-XBP1 pathway promotes T helper cell differentiation by resolving secretory stress and accelerating proliferation. Genome Medicine 10:\newline

Miragaia, RJ; Zhang, X; \textbf{Gomes, T}; Svensson, V; Ilicic, T; Henriksson, J; Kar, G; Lönnberg, T. (2018) Single-cell RNA-sequencing resolves self-antigen expression during mTEC development, Scientific Reports, 8, 1, 685, Nature Publishing Group\newline

